---
title: Modelling in Physical Geography
subtitle: Introduction to R scripting - Day 2
output: ioslides_presentation
---

## Graphical representation of data
```{r, echo = FALSE, fig.height = 5, fig.width = 5, fig.align = "center"}
smap <- MASS::mvrnorm(100, c(0,0), Sigma = matrix(c(1,0.7,0.7,1), ncol = 2))
par(mfrow = c(2,2), mar = c(2,2,3,3))
barplot(table(sample(LETTERS[1:3], size = 100, replace = TRUE, prob = c(1,1.5,0.2))), main = "Barplot", ylab = "n")
boxplot(smap[,1], main = "Boxplot")
hist(smap[,1], main = "Histogram", xlab = "", ylab = "n")
plot(smap, main = "Scatterplot", xlab = "Variable 1", ylab = "Variable 2")
```

# Summary statistics

## Mean
The average value

```{r,echo=FALSE}
skew <- rlnorm(1000,0,0.5)
hist(skew, main = "", xlab = "", breaks = 15)
abline(v=mean(skew), col = "red", lwd = 2, lty = 2)
```

## Median
The middle value

```{r,echo=FALSE}
hist(skew, main = "", xlab = "", breaks = 15)
abline(v=mean(skew), col = "red", lwd = 2, lty = 2)
abline(v=median(skew), col = "blue", lwd = 2, lty = 2)
```

## Standard deviation
A measure of spread

```{r,echo=FALSE}
r <- mean(skew)
hist(skew, main = "", xlab = "", breaks = 15)
abline(v=mean(skew), col = "red", lwd = 2, lty = 2)
abline(v=median(skew), col = "blue", lwd = 2, lty = 2)
segments(x0=r, x1 = c(r+sd(skew),r-sd(skew)), y0 = 350)
```


## Quantiles
Generalized median. For example, the 0.15 quantile is the smallest value that is greater than 15% of observations (also called the 15th percentile).

```{r echo = FALSE}
hist(skew, xlab = "", breaks = 15, main = "")
abline(v=mean(skew), col = "red", lwd = 2, lty = 2)
abline(v=median(skew), col = "blue", lwd = 2, lty = 2)
segments(x0=r, x1 = c(r+sd(skew),r-sd(skew)), y0 = 350)
abline(v = quantile(skew, probs = c(0.15)), lwd = 3, col = "tomato3")
```

## Normal distribution
A continuous distribution parametrized by its mean and standard deviation.

```{r echo = FALSE, fig.width = 7, fig.align = "center"}
plot(dnorm(seq(-4,4,0.05)), type = "l", xlab = "", ylab = "", axes  =FALSE, ylim = c(0,0.5))
lines(dnorm(seq(-4,4,0.05), sd = 2), type = "l", xlab = "", ylab = "", lty = 2)
text(x = 75, y = 0.41, labels = "sd = 1")
text(x = 75, y = 0.21, labels = "sd = 2")
```

## Central limit theorem
If one takes multiple random samples from almost any continuous distribution, the sample means will follow a normal distribution centered on the true mean.

## Scaling variables
Substracting from each member of a sample the sample mean, and then dividing them by the sample standard deviation.

Facilitates comparisons between variables on different scales. A scaled variable has mean zero and a standard deviation of one.

## Bootstrapping
Estimating the uncertainty in a point estimate of a distribution statistic caused by sampling error. Performed by resampling the sample with replacement to aquire the approximate sampling distribution of the statistic of interest.
